---
title: "Download raw files from GEO"
author: "mparilla"
date: "2019-07-03"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Setup

Clear workspace
```{r}
rm(list = ls())
```

Set default to not run chunks when knitting
```{r}
knitr::opts_chunk$set(eval = FALSE)
```

Load packages
```{r}
library("tidyverse")
library("GEOquery")
library("Biobase")
library("affy")
```


## Import Datasets

#### GSE6338

We will use the metadata and clinical annotation from the normalizeed data and then combine this with the raw CEL intensity readings that we normalize ourselves. 

```{r}
# download normalized data with full metadata
gse6338 <- getGEO("GSE6338")
saveRDS(gse6338, "data/gse_6338.rds")

```

```{r}
# load previously downloaded data
gse6338 <- readRDS("data/gse_6338.rds")
str(gse6338)
meta6338 <- phenoData(gse6338[[1]])
meta_data6338 <- pData(meta6338)
class(meta_data6338)
str(meta_data6338)
```

```{r}
# get raw data from gse6338
filePaths = getGEOSuppFiles("GSE6338", makeDirectory = TRUE)
```


#### GSE14879

```{r}
# download normalized data with full metadata
gse14879 <- getGEO("GSE14879")
saveRDS(gse14879, "data/gse_14879.rds")
```

```{r}
# load previously downloaded data
gse14879 <- readRDS("data/gse_14879.rds")
str(gse14879)
meta14879 <- phenoData(gse14879[[1]])
meta_data14879 <- pData(meta14879)
class(meta_data14879)
str(meta_data14879)
```

```{r}
# get raw data from gse14879
filePaths = getGEOSuppFiles("GSE14879", makeDirectory = TRUE)
```


#### GSE19069
```{r}
# download normalized data with full metadata
gse19069 <- getGEO("GSE19069")
str(gse19069)
saveRDS(gse19069, "data/gse_19069.rds")
```

```{r}
# load previously downloaded data
gse19069 <- readRDS("data/gse_19069.rds")
meta19069 <- phenoData(gse19069[[1]])
meta_data19069 <- pData(meta19069)
class(meta_data19069)
str(meta_data19069)
```

```{r}
# get raw data from gse19069
filePaths = getGEOSuppFiles("GSE19069", makeDirectory = TRUE)

```

## Combine CEL files

Using affy package
https://bioconductor.org/packages/release/bioc/vignettes/affy/inst/doc/affy.pdf

Following this tutorial: http://rstudio-pubs-static.s3.amazonaws.com/16793_fba4b435fe4e4c17bf6a13b5d8d05eec.html

```{r}
combined_GSE6338 <- ReadAffy(celfile.path = "data/GSE6338")
combined_GSE19069 <- ReadAffy(celfile.path = "data/GSE19069")
combined_GSE14879 <- ReadAffy(celfile.path = "data/GSE14879")
```

Normalization
```{r}
GSE6338_rma <- rma(combined_GSE6338)
GSE19069_rma <- rma(combined_GSE19069)
GSE14879_rma <- rma(combined_GSE14879)

# data now in the format of normalized expressionSet
```

Expressionset reference: https://www.bioconductor.org/packages/release/bioc/vignettes/Biobase/inst/doc/ExpressionSetIntroduction.pdf

## Data cleaning

Get phenotype data from expressionsets
```{r}
p14879 <- pData(gse14879[[1]])
p6338 <- pData(gse6338[[1]])
p19069 <- pData(gse19069[[1]])
```

```{r}
p14879_fixed <- p14879 %>%
  as_tibble() %>%
  mutate(rnames = geo_accession) %>%
  dplyr::rename(group = title) %>%
  as.data.frame() %>%
  column_to_rownames(var = "rnames")
p14879_fixed[1:5,1:5]
```

Fix column names for expression data
```{r}
exp6338 <- exprs(GSE6338_rma)
exp6338[1:5, 1:5]

exp6338_fixed <- exp6338 %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column(var = "sample_id") %>%
  as_tibble() %>%
  tidyr::separate(sample_id, 
                  into = c("sample_id", NA), 
                  remove = TRUE) %>%
  as.data.frame() %>%
  column_to_rownames(var = "sample_id") %>%
  as.matrix() %>%
  t()

exp6338_fixed[1:5, 1:5]  

#-----------------------------------------
exp19069 <- exprs(GSE19069_rma)
exp19069[1:5, 1:5]

exp19069_fixed <- exp19069 %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column(var = "sample_id") %>%
  as_tibble() %>%
  tidyr::separate(sample_id, 
                  into = c("sample_id", NA), 
                  remove = TRUE) %>%
  as.data.frame() %>%
  column_to_rownames(var = "sample_id") %>%
  as.matrix() %>%
  t()

exp19069_fixed[1:5, 1:5]  

#-----------------------------------------
exp14879 <- exprs(GSE14879_rma)
exp14879[1:5, 1:5]

exp14879_fixed <- exp14879 %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column(var = "sample_id") %>%
  as_tibble() %>%
  tidyr::separate(sample_id, 
                  into = c("sample_id", NA), 
                  remove = TRUE) %>%
  as.data.frame() %>%
  column_to_rownames(var = "sample_id") %>%
  as.matrix() %>%
  t()

exp14879_fixed[1:5, 1:5]  

```

## Assemble into single dataframes

Expression data
```{r}
identical(rownames(exp6338_fixed), rownames(exp14879_fixed))
identical(rownames(exp19069_fixed), rownames(exp14879_fixed))

exp6338_df <- exp6338_fixed %>%
  as.data.frame() %>%
  rownames_to_column(var = "probe_id") %>%
  as_tibble()

exp14879_df <- exp14879_fixed %>%
  as.data.frame() %>%
  rownames_to_column(var = "probe_id") %>%
  as_tibble()

exp19069_df <- exp19069_fixed %>%
  as.data.frame() %>%
  rownames_to_column(var = "probe_id") %>%
  as_tibble()
  
combined_exprs <- exp6338_df %>%
  left_join(exp14879_df, by = "probe_id") %>%
  left_join(exp19069_df, by = "probe_id") %>%
  as.data.frame() %>%
  column_to_rownames(var = "probe_id") %>%
  as.matrix()

combined_exprs[1:5, 1:5]
dim(combined_exprs)
str(combined_exprs)

# save for later use
saveRDS(combined_exprs, "output/combined_exprs.rds")
```

Phenotype data
```{r}
combined_exprs <- readRDS("output/combined_exprs.rds")

p14879 <- pData(gse14879[[1]])
p6338 <- pData(gse6338[[1]])
p19069

p14879_df <- p14879_fixed %>%
  rownames_to_column(var = "sample_id") %>%
  as_tibble() %>%
  print()

p6338_df <- p6338 %>%
  rownames_to_column(var = "sample_id") %>%
  as_tibble() %>%
  dplyr::rename(group = title) %>%
  print()

p19069_df <- p19069 %>%
  rownames_to_column(var = "sample_id") %>%
  as_tibble() %>%
  dplyr::rename(group = title) %>%
  print()

match_table <- colnames(combined_exprs) %>%
  enframe() %>%
  dplyr::select(sample_id = value) %>%
  print()

subtable1 <- match_table %>%
  left_join(p14879_df, by = "sample_id")

subtable2 <- subtable1 %>%
  left_join(p6338_df, by = "sample_id")

combined_pdata <- subtable2 %>%
  left_join(p19069_df, by = "sample_id") %>%
  print()

combined_pdata_groups <- combined_pdata %>%
  mutate(group.x = as.character(group.x),
         group.y = as.character(group.y),
         group = as.character(group)) %>%
  mutate(groups1 = coalesce(group.x, group.y)) %>%
  mutate(groups = coalesce(group, groups1)) %>%
  dplyr::select(sample_id, groups, everything()) %>%
  print()

write_csv(combined_pdata_groups, "output/combined_pdata_prelim_groups.csv")
```


```{r}
combined_pdata_fixed <- read_csv("output/combined_pdata_fixed_groups.csv")

combined_pdata_fixed
summary(as.factor(combined_pdata_fixed$collapsed_group))

pheno_data <- combined_pdata_fixed %>%
  as.data.frame() %>%
  column_to_rownames(var = "sample_id")

pheno_data[1:5, 1:5]

cnames <- colnames(combined_exprs) %>% as.character() %>% print()
rnames <- rownames(pheno_data) %>% as.character() %>% print()
identical(rnames, cnames)
setdiff(rnames, cnames)

```

Taken from documentation for bioconductor expression
```{r}
# create a frame for annotation of column names in metadata
# will be empty for now
varnames <- colnames(pheno_data)
vardata <- data.frame(labelDescription = varnames, row.names = varnames)

# build an AnnotatedDataFrame
tcl_phenoData <- new("AnnotatedDataFrame", 
                     data = pheno_data,
                     varMetadata = vardata)

# convert expression data to matrix
# already done

# make annotation

anno = "Three datasets downloaded from GEO: GSE6338, GSE14879, GSE19069. Then downloaded raw CEL files and normalized, replaced expression data with our own normalized data and refactored phenotype data with groups corresponding to lymphoma type or control."

tcl_eset <- ExpressionSet(assayData = combined_exprs,
                          phenoData = tcl_phenoData,
                          annotation = anno)

saveRDS(tcl_eset, "output/tcl_eset.rds")

```
