---
title: "gsva"
author: "mleukam"
date: "2019-09-20"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Setup

clear the workspace
```{r}
rm(list = ls())
```

Set default to not run chunks when knitting
(long operations in notebook make updating html output time prohibitive)
```{r}
knitr::opts_chunk$set(eval = FALSE)
```

Read in packages
```{r}
library(tidyverse)
library(edgeR)
library(limma)
library(msigdbr)
library(GSA)
library(biomaRt)
library(Biobase)
library(GSVA)
library(ggsci)
library(ggpubr)
library(ggbeeswarm)
library(broom)
```

Load data
```{r}
tcl_eset <- readRDS("output/tcl_eset.rds")
```

## Prepare gene sets

http://software.broadinstitute.org/gsea/msigdb/search.jsp

```{r}
gsetnames <- c("BAKKER_FOXO3_TARGETS_UP", "DELPUECH_FOXO3_TARGETS_UP", "GSE21678_WT_VS_FOXO1_FOXO3_KO_TREG_UP", "FOXO3_01")

# use msigdbr to retreive human gene sets from MSigDB
m_df = msigdbr(species = "Homo sapiens")

# review subcategories
summary(as.factor(m_df$gs_subcat))

# select gene sets from msigdbr-retreived gene list with vector of gene set names
my_gsets <- m_df %>%
  dplyr::filter(gs_name %in% gsetnames)
summary(as.factor(my_gsets$gs_name))

# make a list of gene symbol vectors for each gene set
gsets_msig <- my_gsets %>%
  dplyr::select(gs_name, human_gene_symbol) %>%
  nest(key = "human_gene_symbol") %>%
  print()
gsets_msig_list <- map(gsets_msig, as.list)
gsets_vectors <- gsets_msig_list$data %>%
  map(pull)
names(gsets_vectors) <- gsets_msig %>% pull(gs_name)
gsets_vectors
str(gsets_vectors)
```

## Prepare expression data
```{r}
expr_matrix <- exprs(tcl_eset)
expr_matrix[1:5, 1:5]
```

Convert affymetrix probe IDs to gene symbols
```{r}
mart <- useMart("ENSEMBL_MART_ENSEMBL")
mart <- useDataset("hsapiens_gene_ensembl", mart)
annotLookup <- getBM(
  mart = mart,
  attributes = c(
    "affy_hg_u133_plus_2",
    "ensembl_gene_id",
    "gene_biotype",
    "external_gene_name"),
  filter = "affy_hg_u133_plus_2",
  values = rownames(exprs(tcl_eset)),
  uniqueRows = TRUE)
```

##Filtering 

#### Valid gene names
```{r}
# apply new rownames
# keep only one entry per probe name
new_rnames <- rownames(expr_matrix) %>%
  enframe() %>%
  dplyr::select(affy_hg_u133_plus_2 = value) %>%
  left_join(annotLookup) %>%
  distinct(affy_hg_u133_plus_2, .keep_all = TRUE) %>%
  pull(external_gene_name) 
gnames_expr_matrix <- expr_matrix
rownames(gnames_expr_matrix) <- new_rnames

# remove "NA" gene names
dim(gnames_expr_matrix)
gnames_expr_matrix_nona <- gnames_expr_matrix %>%
  as.data.frame() %>%
  rownames_to_column(var = "gene_name") %>%
  filter(gene_name != is.na(gene_name)) %>%
  column_to_rownames(var = "gene_name") %>%
  as.matrix()
gnames_expr_matrix_nona[1:5, 1:5]
dim(gnames_expr_matrix_nona)

# collapse identical gene names by averaging
dim(gnames_expr_matrix_nona)
expr_matrix_collapsed <- avereps(gnames_expr_matrix_nona)
dim(expr_matrix_collapsed)
```

#### Protein Coding
```{r}
summary(as.factor(annotLookup$gene_biotype))

dim(expr_matrix_collapsed)
expr_matrix_prtcode <- expr_matrix_collapsed %>%
  as.data.frame() %>%
  rownames_to_column(var = "external_gene_name") %>%
  as_tibble() %>%
  left_join(annotLookup) %>%
  distinct(external_gene_name, .keep_all = TRUE) %>%
  dplyr::filter(gene_biotype == "protein_coding") %>%
  dplyr::select(-affy_hg_u133_plus_2, -ensembl_gene_id, -gene_biotype) %>%
  as.data.frame() %>%
  column_to_rownames(var = "external_gene_name") %>%
  as.matrix()
dim(expr_matrix_prtcode)
```

#### Expression level
```{r}
pheno_data <- pData(tcl_eset)
summary(as.factor(pheno_data$collapsed_group))

median(expr_matrix_prtcode)
mean(expr_matrix_prtcode)

# filter out genes that aren't at least expressed greater than 1 in at least half of all cases
total_stats <- data.frame(
  total = apply(expr_matrix_prtcode, 1, 
                function(x) sum(x > 1, 
                                na.rm = TRUE)))
keep <- which(total_stats$total >= 135) 

dim(expr_matrix_prtcode)
expr_matrix_filtered <- expr_matrix_prtcode[keep,]
dim(expr_matrix_filtered)
```

#### Density Plot
```{r}
# Density plots
# tidy data
tidy_expr <- expr_matrix_filtered %>%
  as.data.frame() %>%
  rownames_to_column(var = "gene_name") %>%
  as_tibble() %>%
  gather(key = "sampleID", value = "intensity", -gene_name) %>%
  print()

# plot
dplot_expr <- ggplot(tidy_expr, aes(intensity)) +
  geom_density() + 
  theme(legend.position = "none")
dplot_expr
```

## Normalize gene expression distributions

Normalisation by the method of trimmed mean of M-values (TMM) is performed using the calcNormFactors function in edgeR. The normalisation factors calculated here are used as a scaling factor for the library sizes. 

```{r}
# get normalization factors
norm_factors <- calcNormFactors(expr_matrix_filtered, method = "TMM")

# apply factor to each column
expr_matrix_norm <- map2_dfc(as.data.frame(expr_matrix_filtered),
                             norm_factors, `*`) %>%
  as.matrix()
rownames(expr_matrix_norm) <- rownames(expr_matrix_filtered)
expr_matrix_norm[1:5, 1:5]
```

## GSVA
Write out matrix and gene set lists
```{r}
# gset list
saveRDS(gsets_vectors, "output/gsets_4.rds")

expr_matrix <- expr_matrix_norm %>% 
  as.data.frame() %>%
  rownames_to_column(var = "gene")

write_csv(expr_matrix, "output/expr_matrix.csv")
```

GSVA
```{r}
tcl_es <- gsva(expr_matrix_norm, gsets_vectors,
                annotation = NULL,
                method = "gsva",
                mx.diff = TRUE,
                verbose = TRUE)

```

GSVA results
```{r}
# pull out group membership
group_lookup <- pheno_data %>%
  rownames_to_column(var = "sample_id") %>%
  as_tibble() %>%
  dplyr::select(sample_id, collapsed_group) %>%
  print()

# add group membership to gsva score matrix
tcl_gsva_tidy <- tcl_es %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column(var = "sample_id") %>%
  as_tibble() %>%
  left_join(group_lookup, by = "sample_id") %>%
  dplyr::select(sample_id, disease = collapsed_group, everything()) %>%
  # filter for relevant groups
  dplyr::filter(disease %in% c("AILD", "ALCL", "control", "PTCL")) %>%
  mutate(disease = as.factor(disease)) %>%
  gather(key = gene_set, value = gsva_score, -sample_id, -disease) %>%
  mutate(gene_set = as.factor(gene_set),
         gene_set = fct_recode(gene_set,
                               "Bakker FOXO3 targets" = "BAKKER_FOXO3_TARGETS_UP",
                               "Delpuech FOXO3 targets" = "DELPUECH_FOXO3_TARGETS_UP",
                               "Predicted FOXO3 targets" = "FOXO3_01",
                               "Genes expressed in wt \nbut not FOXO1_FOXO3 KO" = "GSE21678_WT_VS_FOXO1_FOXO3_KO_TREG_UP")) %>%
  group_by(gene_set) %>%
  print()
  
# plot gsva scores in boxplots
plot1 <- ggplot(data = tcl_gsva_tidy, aes(x = gene_set, y = gsva_score, fill = disease)) +
  geom_boxplot() +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 40, hjust = 1)) +
  geom_hline(yintercept = 0, linetype = 3) +
  scale_fill_nejm(alpha = 0.7) +
  ggtitle("Gene Set Expression by Individual Samples per Disease Group") +
  xlab("Gene Set") +
  ylab("GSVA Score") +
  labs(fill = "Disease")

plot1
```

Test results for significance
```{r}
tcl_gsva_tidy

# compute one-way anova test for each gene set to see if there is any significant difference among means
compare_means(gsva_score ~ disease, 
              data = tcl_gsva_tidy,
              method = "anova", 
              group.by = "gene_set")

# find out which pairwise comparisons are significant
# adjust for multiple hypothesis testing!
tcl_gsva_bakker <- tcl_gsva_tidy %>%
  dplyr::filter(gene_set == "Bakker FOXO3 targets")
mod_bakker <- aov(gsva_score ~ disease, data = tcl_gsva_bakker) 
summary(mod_bakker)
bakker <- tidy(TukeyHSD(mod_bakker)) %>% 
  bind_cols(gene_set = rep("Bakker", 6)) %>%
  dplyr::select(gene_set, comparison, adj.p.value, everything()) %>%
  print()

tcl_gsva_delpuech <- tcl_gsva_tidy %>%
  dplyr::filter(gene_set == "Delpuech FOXO3 targets")
mod_delpuech <- aov(gsva_score ~ disease, data = tcl_gsva_delpuech) 
summary(mod_delpuech)
delpuech <- tidy(TukeyHSD(mod_delpuech)) %>% 
  bind_cols(gene_set = rep("Delpuech", 6)) %>%
  dplyr::select(gene_set, comparison, adj.p.value, everything()) %>%
  print()

tcl_gsva_predicted <- tcl_gsva_tidy %>%
  dplyr::filter(gene_set == "Predicted FOXO3 targets")
mod_predicted <- aov(gsva_score ~ disease, data = tcl_gsva_predicted) 
summary(mod_predicted)
predicted <- tidy(TukeyHSD(mod_predicted)) %>% 
  bind_cols(gene_set = rep("predicted", 6)) %>%
  dplyr::select(gene_set, comparison, adj.p.value, everything()) %>%
  print()

tcl_gsva_ko <- tcl_gsva_tidy %>%
  dplyr::filter(gene_set == "Genes expressed in wt \nbut not FOXO1_FOXO3 KO")
mod_ko <- aov(gsva_score ~ disease, data = tcl_gsva_ko) 
summary(mod_ko)
KO <- tidy(TukeyHSD(mod_ko)) %>% 
  bind_cols(gene_set = rep("KO", 6)) %>%
  dplyr::select(gene_set, comparison, adj.p.value, everything()) %>%
  print()

pairwise_comparisons <- bakker %>%
  bind_rows(delpuech, predicted, KO) %>%
  arrange(comparison) %>%
  print(n = 24)
```

